---
title: "CMGD Parquet Play"
format: html
---

# Background

We have created parquet files from the CMGD data. 
This is a demonstration of how to read the parquet files into R
using duckdb. Three tables exist as parquet files living in 
cloud storage. The tables are:

- Marker abundances
- Marker presences
- Metaphlan bugs list

# Setup

Install duckdb, DBI, dbplyr, and dplyr.

Then, load the libraries.

```{r}
library(duckdb)
library(DBI)
library(dbplyr)
library(dplyr)
```

# Connect to the database

```{r}
con <- dbConnect(duckdb::duckdb(), ":memory:")
```

And set up the ability to read from cloud storage using duckdb:

```{r}
dbExecute(con, "install httpfs;")
dbExecute(con, "load httpfs;")
```

# Create views

Views are created to make it easier to work with the data. They do not make 
a copy of the data, but instead just provide a way to access the data in a
more convenient way.

These next few lines will take a while to run because they are reading the
parquet files from cloud storage. Subsequent runs will often be faster
because the metadata will be cached. 

```{r}
dbExecute(con, paste(
  "create view metaphlan_bugs as select * from",
  "read_parquet('https://store.cancerdatasci.org/cmgd/cMDv4/metaphlan_bugs.parquet');"
))
```

We can check to see that this works by running a query:

```{r}
dbListTables(con)
```

# Query the data

Now that we have the data loaded, we can query it. 

```{r}
dbGetQuery(con, "select * from metaphlan_bugs limit 5;")
```
Any SQL query can be run on the data.

As an alternative, we can use dplyr to query the data. 

```{r}
metaphlan_bugs <- tbl(con, "metaphlan_bugs")
metaphlan_bugs |>
  select(sample_id) |>
  group_by(sample_id) |>
  summarise(bug_count = n()) |>
  arrange(desc(bug_count)) |>
  head(10)
```

As another example, we can get the metaphlan results for a specific sample:

```{r}
metaphlan_bugs |>
  filter(sample_id == "SAMN02334067") |>
  select(sample_id, clade_name, relative_abundance) |>
  arrange(desc(relative_abundance)) -> metaphlan_bugs_sample

head(metaphlan_bugs_sample, 50)
```

You can create tables for the other two parquet files in the same way. The 
urls of the other two files are:

- https://store.cancerdatasci.org/cmgd/cMDv4/marker_abundances.parquet
- https://store.cancerdatasci.org/cmgd/cMDv4/marker_presences.parquet

Those files are larger, so working with them will take a little longer, but
the process is the same.

If you want to work with the parquet files locally, you can download them
and change the URLs in the `read_parquet` calls to point to the local files.
This could be faster, but it will require disk space. 

Finally, you can insert the data into a local database if you want to work
directly with duckdb tables rather than views over parquet files. The speed 
may increase a bit, but not by a LOT. 

